---
title: "Simulating and interpreting SIR dynamics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

In this walk through, we will be simulating an SIR model.

The SIR model is a simple compartmental model used to describe the spread of infectious diseases in a population. The model divides the population into three compartments: susceptible (S), infectious (I), and recovered (R) individuals. The model equations describe the flow of individuals between these compartments based on the transmission rate ($\beta$) and the recovery rate ($\gamma$).

<!-- ```{r, child=c('./scripts/rmd_child_01_lsoda_explanation.Rmd')} -->
<!-- ``` -->

We will be using the `{deSolve}` package to solve the model equations. The `{deSolve}` package is a powerful and flexible package for solving differential equations in R. It provides a wide range of solvers for different types of differential equations.

Here, we will use the `lsoda()` function from the `{deSolve}` package to solve the SIR model. The `lsoda()` function is a wrapper for the Livermore Solver for Ordinary Differential Equations (LSODA) developed by Hindmarsh and Petzold. It is a robust and efficient solver for stiff and non-stiff differential equations.

The `lsoda()` function takes, among others, the following arguments: - `func`: the model function that defines the differential equations, - `times`: the time points at which to return results, - `y`: the initial conditions, and - `parms`: the parameters of the model.

## The model function

Let's start by defining the model equations.

Recall that the model equations for the SIR model are given by:

\begin{align}
\frac{dS}{dt} & = \color{orange}{-\beta S I} \\
\frac{dI}{dt} & = \color{orange}{\beta S I} - \color{blue}{\gamma I} \\
\frac{dR}{dt} & = \color{blue}{\gamma I}
\end{align}

where:

- $S$ is the number of susceptible individuals,
- $I$ is the number of infectious individuals,
- $R$ is the number of recovered individuals,
- $\beta$ is the transmission rate,
- $\gamma$ is the recovery rate.


The model function in R is defined as follows:

```{r sir-model}
# 1. Define the SIR model
sir_model <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    return(list(c(dS, dI, dR)))
  })
}
```

<!-- ```{r, child=c('./scripts/rmd_child_02_model_function_components.Rmd')} -->
<!-- ``` -->

Note the following about the function defined above:

-   We use the `with()` function to evaluate the expressions in the context of the list created by `as.list(c(state, parameters))`. This allows us to access the state variables and parameters directly without having to use the `$` operator. This makes the code easier to read and write (Run `?with` in the R console and read the documentation of the first three arguments).

-   The function returns a list of the derivatives of the state variables. This is the format expected by the `lsoda()` function (Run `?deSolve::lsoda` in the R console and read the documentation of the first three arguments).

## Model parameters

Next, we will define the parameter values and initial conditions.

The model has two parameters: $\beta$ and $\gamma$.

$\beta$ is a composite of different terms and is not intuitive at face value, so we will define it in terms of the basic reproduction number ($R0$) and the infectious period using the relation $R0 = \frac{\beta}{\gamma}$.

```{r sir-model-params}
# 2. Define the parameter 

R0 <- 2

# The infectious period is the average duration for which an individual remains infectious.
infectious_period <- 14
# Remember gamma = 1/ infectious_period as discussed in the slides
gamma <- 1/infectious_period 

# The define beta, we will use the relation R0 = beta / gamma  because it is easier to interpret, that is, given an R0 and infectious period, we can calculate the corresponding beta value.
# beta is not directly interpretable because it is a composite of a number of factors.
beta <- R0 * gamma

# The parameters
params <- c(
  beta = beta,
  gamma = gamma
)
```


## Initial conditions

Due to numerical stability issues, we will specify the populations as proportions and assume they sum up to $1$. We will then scale the populations to the actual population size after obtaining the results.

We will assume the outbreak starts with $0.01$ infectious individual at the start of the simulation. The population is completely susceptible at the start of the simulation, so the number of susceptible individuals is $N - 0.01$. The number of recovered individuals is zero at the start of the simulation.
```{r sir-model-init}
#3. Define the initial conditions
N  <- 1
I0 <- 0.01
# Initial conditions for S, I, R
inits <- c(S = N - I0, I = I0, R = 0)
```

## Time steps to return results

`lsoda()` also expects a vector of time points at which to return results. We will run the model for 365 days and return results at each day.
```{r sir-time}
# 4. Define the time points to return results
dt <- 1:365
```

## Solving the model

Now, we are ready to solve the model. We will solve the model using the `lsoda()` function from the `{deSolve}` package. For now, we will use the default values of the function. You are encouraged to explore the documentation to understand the function better.
```{r sir-model-solve}
#5. Solve the model
# Load the necessary libraries
library(deSolve)
# Solve the model
results <- deSolve::lsoda(
  y = inits,
  times = dt,
  func = sir_model,
  parms = params
)
```

The results are stored in a matrix with the first column representing the time points and the subsequent columns representing the number of individuals in the susceptible, infectious, and recovered compartments at each time point.

To work with the results more easily, we will convert the results to a data frame.

```{r sir-model-results}
#6. Manipulate and interpret the results
# Make it a data.frame
results <- as.data.frame(results)
head(results)
```

### Plotting the results

Now, let's plot the results.

We will use the `ggplot2` package to plot the results. We will reshape the data to long format to make it easier to plot. For this, we will use the `pivot_longer()` function from the `tidyr` package.
```{r sir-model-plot}
# Load the necessary libraries
library(ggplot2)
library(tidyr)
# Create data for ggplot2 by reshaping
results_long <- results |>
  pivot_longer(
    cols = c(2:4),
    names_to = "compartment",
    values_to = "value"
  )

sir_plot <- ggplot(
  data = results_long,
  aes(
    x = time,
    y = value,
    color = compartment
  )
  ) +
  geom_line(linewidth = 1) +
  labs(
    title = "SIR model",
    x = "Time",
    y = "Number of individuals"
  )
plot(sir_plot)
```
```{r}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# 1. Define the SIR model
sir_model <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    return(list(c(dS, dI, dR)))
  })
}

# 2. Define the parameter 

R0 <- 3

# The infectious period is the average duration for which an individual remains infectious.
infectious_period <- 14
# Remember gamma = 1/ infectious_period as discussed in the slides
gamma <- 1/infectious_period 

# The define beta, we will use the relation R0 = beta / gamma  because it is easier to interpret, that is, given an R0 and infectious period, we can calculate the corresponding beta value.
# beta is not directly interpretable because it is a composite of a number of factors.
beta <- R0 * gamma

# The parameters
params <- c(
  beta = beta,
  gamma = gamma
)

#3. Define the initial conditions
N  <- 1
I0 <- 0.01
# Initial conditions for S, I, R

# 4. Define the time points to return results
dt <- 1:365

#5. Solve the model
# Load the necessary libraries
library(deSolve)
# Solve the model
results <- deSolve::lsoda(
  y = inits,
  times = dt,
  func = sir_model,
  parms = params
)

#6. Manipulate and interpret the results
# Make it a data.frame
results <- as.data.frame(results)
head(results)

### Plotting the results

# Load the necessary libraries
library(ggplot2)
library(tidyr)
# Create data for ggplot2 by reshaping
results_long <- results |>
  pivot_longer(
    cols = c(2:4),
    names_to = "compartment",
    values_to = "value"
  )

sir_plot1 <- ggplot(
  data = results_long,
  aes(
    x = time,
    y = value,
    color = compartment
  )
  ) +
  geom_line(linewidth = 1) +
  labs(
    title = "SIR model",
    x = "Time",
    y = "Number of individuals"
  )
plot(sir_plot1)
```

## R0 = c(1, 2, 3, 4, 5)
```{r}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# 1. Define the SIR model
sir_model <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    return(list(c(dS, dI, dR)))
  })
}

# 2. Define the parameter 

# List of R0 values to simulate
R0_values <- c(1, 2, 3, 4, 5)

# The infectious period is the average duration for which an individual remains infectious.
infectious_period <- 14
gamma <- 1 / infectious_period

# 3. Define the initial conditions
N  <- 1
I0 <- 0.01
S0 <- N - I0
R0_init <- 0
inits <- c(S = S0, I = I0, R = R0_init)

# 4. Define the time points to return results
dt <- 1:365

# Load the necessary libraries
library(deSolve)
library(ggplot2)
library(tidyr)

# 5. Solve the model for each R0 and store the results
results_list <- lapply(R0_values, function(R0) {
  beta <- R0 * gamma
  params <- c(beta = beta, gamma = gamma)
  results <- deSolve::lsoda(y = inits, times = dt, func = sir_model, parms = params)
  results <- as.data.frame(results)
  results$R0 <- R0
  return(results)
})

# Combine all results into one data frame
all_results <- do.call(rbind, results_list)

# 6. Manipulate and interpret the results
# Create data for ggplot2 by reshaping
results_long <- all_results |>
  pivot_longer(
    cols = c(S, I, R),
    names_to = "compartment",
    values_to = "value"
  )

# Plot the results
sir_plot2 <- ggplot(
  data = results_long,
  aes(
    x = time,
    y = value,
    color = interaction(compartment, R0),
    group = interaction(compartment, R0)
  )
) +
  geom_line(linewidth = 1) +
  labs(
    title = "SIR Model for Different R0 Values",
    x = "Time",
    y = "Number of Individuals",
    color = "Compartment.R0"
  ) +
  theme_minimal()

# Display the plot
print(sir_plot2)
```


```{r}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# 1. Define the SIR model
sir_model <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    return(list(c(dS, dI, dR)))
  })
}

# 2. Define the parameter 

R0 <- 2

# The infectious period is the average duration for which an individual remains infectious.
infectious_period <- 3
# Remember gamma = 1/ infectious_period as discussed in the slides
gamma <- 1/infectious_period 

# The define beta, we will use the relation R0 = beta / gamma  because it is easier to interpret, that is, given an R0 and infectious period, we can calculate the corresponding beta value.
# beta is not directly interpretable because it is a composite of a number of factors.
beta <- R0 * gamma

# The parameters
params <- c(
  beta = beta,
  gamma = gamma
)

#3. Define the initial conditions
N  <- 1
I0 <- 0.01
# Initial conditions for S, I, R

# 4. Define the time points to return results
dt <- 1:365

#5. Solve the model
# Load the necessary libraries
library(deSolve)
# Solve the model
results <- deSolve::lsoda(
  y = inits,
  times = dt,
  func = sir_model,
  parms = params
)

#6. Manipulate and interpret the results
# Make it a data.frame
results <- as.data.frame(results)
head(results)

### Plotting the results

# Load the necessary libraries
library(ggplot2)
library(tidyr)
# Create data for ggplot2 by reshaping
results_long <- results |>
  pivot_longer(
    cols = c(2:4),
    names_to = "compartment",
    values_to = "value"
  )

sir_plot3 <- ggplot(
  data = results_long,
  aes(
    x = time,
    y = value,
    color = compartment
  )
  ) +
  geom_line(linewidth = 1) +
  labs(
    title = "SIR model",
    x = "Time",
    y = "Number of individuals"
  )
plot(sir_plot3)
```
```{r}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# 1. Define the SIR model
sir_model <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    return(list(c(dS, dI, dR)))
  })
}

# 2. Define the parameter 

R0 <- 2

# The infectious period is the average duration for which an individual remains infectious.
infectious_period <- 10
# Remember gamma = 1/ infectious_period as discussed in the slides
gamma <- 1/infectious_period 

# The define beta, we will use the relation R0 = beta / gamma  because it is easier to interpret, that is, given an R0 and infectious period, we can calculate the corresponding beta value.
# beta is not directly interpretable because it is a composite of a number of factors.
beta <- R0 * gamma

# The parameters
params <- c(
  beta = beta,
  gamma = gamma
)

#3. Define the initial conditions
N  <- 1
I0 <- 0.01
# Initial conditions for S, I, R

# 4. Define the time points to return results
dt <- 1:365

#5. Solve the model
# Load the necessary libraries
library(deSolve)
# Solve the model
results <- deSolve::lsoda(
  y = inits,
  times = dt,
  func = sir_model,
  parms = params
)

#6. Manipulate and interpret the results
# Make it a data.frame
results <- as.data.frame(results)
head(results)

### Plotting the results

# Load the necessary libraries
library(ggplot2)
library(tidyr)
# Create data for ggplot2 by reshaping
results_long <- results |>
  pivot_longer(
    cols = c(2:4),
    names_to = "compartment",
    values_to = "value"
  )

sir_plot4 <- ggplot(
  data = results_long,
  aes(
    x = time,
    y = value,
    color = compartment
  )
  ) +
  geom_line(linewidth = 1) +
  labs(
    title = "SIR model",
    x = "Time",
    y = "Number of individuals"
  )
plot(sir_plot4)
```

```{r}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# 1. Define the SIR model
sir_model <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    return(list(c(dS, dI, dR)))
  })
}

# 2. Define the parameter 

R0 <- 5

# The infectious period is the average duration for which an individual remains infectious.
infectious_period <- 14
# Remember gamma = 1/ infectious_period as discussed in the slides
gamma <- 1/infectious_period 

# The define beta, we will use the relation R0 = beta / gamma  because it is easier to interpret, that is, given an R0 and infectious period, we can calculate the corresponding beta value.
# beta is not directly interpretable because it is a composite of a number of factors.
beta <- R0 * gamma

# The parameters
params <- c(
  beta = beta,
  gamma = gamma
)

#3. Define the initial conditions
N  <- 1
I0 <- 0.01
# Initial conditions for S, I, R

# 4. Define the time points to return results
dt <- 1:365

#5. Solve the model
# Load the necessary libraries
library(deSolve)
# Solve the model
results <- deSolve::lsoda(
  y = inits,
  times = dt,
  func = sir_model,
  parms = params
)

#6. Manipulate and interpret the results
# Make it a data.frame
results <- as.data.frame(results)
head(results)

### Plotting the results

# Load the necessary libraries
library(ggplot2)
library(tidyr)
# Create data for ggplot2 by reshaping
results_long <- results |>
  pivot_longer(
    cols = c(2:4),
    names_to = "compartment",
    values_to = "value"
  )

sir_plot5 <- ggplot(
  data = results_long,
  aes(
    x = time,
    y = value,
    color = compartment
  )
  ) +
  geom_line(linewidth = 1) +
  labs(
    title = "SIR model",
    x = "Time",
    y = "Number of individuals"
  )
plot(sir_plot5)
```
```{r}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# 1. Define the SIR model
sir_model <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    return(list(c(dS, dI, dR)))
  })
}

# 2. Define the parameter 

R0 <- 1

# The infectious period is the average duration for which an individual remains infectious.
infectious_period <- 14
# Remember gamma = 1/ infectious_period as discussed in the slides
gamma <- 1/infectious_period 

# The define beta, we will use the relation R0 = beta / gamma  because it is easier to interpret, that is, given an R0 and infectious period, we can calculate the corresponding beta value.
# beta is not directly interpretable because it is a composite of a number of factors.
beta <- R0 * gamma

# The parameters
params <- c(
  beta = beta,
  gamma = gamma
)

#3. Define the initial conditions
N  <- 1
I0 <- 0.01
# Initial conditions for S, I, R

# 4. Define the time points to return results
dt <- 1:365

#5. Solve the model
# Load the necessary libraries
library(deSolve)
# Solve the model
results <- deSolve::lsoda(
  y = inits,
  times = dt,
  func = sir_model,
  parms = params
)

#6. Manipulate and interpret the results
# Make it a data.frame
results <- as.data.frame(results)
head(results)

### Plotting the results

# Load the necessary libraries
library(ggplot2)
library(tidyr)
# Create data for ggplot2 by reshaping
results_long <- results |>
  pivot_longer(
    cols = c(2:4),
    names_to = "compartment",
    values_to = "value"
  )

sir_plot6 <- ggplot(
  data = results_long,
  aes(
    x = time,
    y = value,
    color = compartment
  )
  ) +
  geom_line(linewidth = 1) +
  labs(
    title = "SIR model",
    x = "Time",
    y = "Number of individuals"
  )
plot(sir_plot6)
```

## Exercise

Answer the following questions by adding code chunks and text chunks.


1. What do you observe from the plot above? Can you describe the dynamics of the SIR model?
```{r}
plot(sir_plot1)
```
**At day 1, All in the population was susceptible to the disease, 1 possibly infected person, and no recovered person in the population.
As the number of infected started to increase, the susceptible was reducing and the recovered also started increasing. At aroung day 70, the infection time reached a peak and began to fall. Those with the disease mostly recovered after day 100, with 20% of the population not getting the disease. After day 150, the in disease  was no spreading more in the population**

2. Change the value of the basic reproduction number ($R0$) to 3 and observe the effect on the dynamics of the SIR model.

** R0 = 3, infectious days =14**
```{r R0=3}
plot(sir_plot1)
```
** 

3. Plot the I compartments for different values of $R0$ (e.g., 1, 2, 3, 4, 5) on the same plot. What do you observe?
** R0 = 4 **
```{r R0 = 4}
plot(sir_plot2)
```


4. Change the value of the infectious period to 3 and 10 days and observe the effect on the dynamics of infection.
** Infectious days = 3, R0=2 **
```{r}
plot(sir_plot3)
```
** infectious days = 10, R0=2 **
```{r}
plot(sir_plot4)
```


5. What are the features of a disease that dies out quickly assuming $R0 > 1$?
** Infectious days = 14, r0 = 5 **
```{r}
plot(sir_plot5)
```


** Infectious days = 14, r0 = 1 **
```{r}
plot(sir_plot6)
```

